Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-02-05 analysis/paper/paper.Rmd
minor edits on the paper

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could result in far-reaching economic or cultural impacts on local societies. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP. and to western Yunnan before 4600 BP. [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010; @Dal2022]. Therefore, the Neolithic Hengduan Mountain areas are key to understanding the processes of agriculture expansion. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the introduction of cultivated millet and rice simultaneously during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022]. However, how and to what extent the two crops were incorporated into the local cultures is still not fully understood.

@@

<!--Yanyuan area and Guijiabao site-->
We focus on Guijiabao (ca. 4800-4300 BP) located in the Yanyuan Basin in Sichuan Province (Figure \@ref(fig:SW-china-map)). Yanyuan Basin is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges. Guijiabao is especially unique due to its location as the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau, with a continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site with long-term occupation [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (ca. 5000-3700 BP), the Bronze Age (ca. 3200-2700 BP), and the late North Song to early Ming Dynasties (ca. 1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022]. For the Neolithic period, it can be further separated into two subphases, 5000-4500 ca. BP and 4500-3700 ca. BP, according to the content of artifact assemblages and ceramic forms and styles.

<!--the major finding at GJB-->
The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins. Most ceramics have surface decorations, such as incised geometric patterns and impressed dot patterns that resemble ceramic decorations found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan [@Guijiabao2017a]. In addition, cord-marked pottery vessels are also commonly found in the early period. About the plant remains, the flotation analysis in 2015 reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on samples collected from ash pits and stratigraphic layers [@Guijiabao2016]. The earliest domesticated millet grain can be traced back to ca. 5000 BP according to its radiocarbon date [@Hao2022]. A smaller number of rice grains (*Oryza sativa*) have also been found in ash pits where the earliest one dates to 4700 ca. BP. [@Hao2022]. Later, a detailed phytolith analysis for soil samples dating to the Neolithic period (5000 ca. BP to 3700 ca. BP) further suggests the presence of rice phytoliths throughout the temporal sequences, indicative of a mixed millet/rice cultivation prior to 5000 ca. BP. [@Huan2022].

@@

Although the current finding at Guijiabao and Gaoshan support the spread of mixed farming of millet/rice into southwest China around 5000 BP., what their proportions are in local culinary practices during the early time are unclear. For example, the analysis of plant remains and phytolith analysis for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice in neighboring sites, such as Gaoshan and Baodun, Henglanshan, and Baiyangcun in Sichuan and Yunnan [@Hao2022; @Guedes2013]. The analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, but evidence directly reflects food preparation or processing using ceramics still lacks. This study aims to provide direct evidence from food residues in ceramics to investigate culinary practices associated with the cultivation of new crops.

@@

## Samples
We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX). Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and one fragment was from the 2016 season. Those potsherds can be divided into two temporal phases, the early Neolithic period (5000-4500 BP.) and the historical period (1000 BP.). The common ceramic is cord-marked sand-tempered vessels unearthed at Guijiabao, and their forms include jars, urns, bowls, and basins [@Guijiabao2017a; @Hao2022]. Our samples are mostly reddish-brown jar fragments with medium-size sand temper and some of them have incised cord-marked decoration. Although the original function of those jars is uncertain, they should be supposed to be used for cooking or storage according to the char on the surfaces and their archaeological contexts from a settlement [@Hao2022; @Guijiabao2017a]. This research also attempts to confirm that those ceramics were cooking pots used in the daily life of the local residents.

@@

# Discussion
<!--A brief summary of interesting results-->
Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might originate from C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots. The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020]. But we also found a base sample dating to ca. 4500 BP. contains a rich composition of long-chain fatty acids, reflecting a good preservation. We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes. Also, the lack of local modern references for animal fats and plants has impacts on our explanation. However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence at Guijiabao.

@@

<!--discussion about possible ruminant fat origin-->
Ruminant fats identified in the Neolithic potsherds around ca. 4500 BP. from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, accompanied with stone tools for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. But whether horse was part of local food choices in the Neolithic period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. The evidence from Guijuabao also confirms a later presence of horse remains. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats fed on C3 diet based on global references [@Evershed2002]. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016]. A detailed analysis of faunal remains from Guijiabao is still in progress that can provide further information.

We observed a mixture of ruminant fats and terrestrial plants based on a number of Guijiabao potsherds dating to the historical period (ca. 1000 BC.). However, we are unable to determine whether there is a shift in diet over time due to our limited sample size. But the overall depleted δ^13^C values for most potsherds suggest C3 plants might be an important diet for local animals and could also be included in the local diet in Yanyuan during the Neolithic transition. This corresponds with the timing when new crops spread into southwest China. Despite of no C4 plant origin in our samples, one potshard implies a source of C4 plant-fed ruminant. In general, the food source detected from our samples comprised mainly of ruminant fats that are more likely from wildlife. This might be related to local ecological zones and environments where wild animals were easier to obtain. Since zooarchaeological analysis and other studies on stone tools are still in progress, we could only rely on preliminary excavation reports with limited details about associated artifacts. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in food choices during the Neolithic transition.

>>>>

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could result in far-reaching economic or cultural impacts on local societies. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP and to western Yunnan before 4600 BP [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010; @Dal2022]. Therefore, the Neolithic Hengduan Mountain areas are key to understanding the processes of agriculture expansion. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the introduction of cultivated millet and rice simultaneously during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022]. However, how and to what extent the two crops were incorporated into the local cultures is still not fully understood.

@@

<!--Yanyuan area and Guijiabao site-->
We focus on Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)). Guijiabao is especially unique due to its location as the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau, with a continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming Dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022]. For the Neolithic period, it can be further separated into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles.

<!--the major finding at GJB-->
The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins. For early Neolithic, ceramics have surface decorations, such as incised geometric patterns and impressed dot patterns that resemble ceramic decorations found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan [@Guijiabao2017a]. In addition, cord-marked vessels are also found. About the plant remains, the flotation analysis in 2015 reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on samples collected from ash pits and stratigraphic layers [@Guijiabao2016]. The earliest domesticated millet grain can be traced back to 5000 BP according to its radiocarbon date [@Hao2022]. A smaller number of rice grains (*Oryza sativa*) have also been found in ash pits where the earliest one dates to 4700 BP [@Hao2022]. Later, a detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences, indicative of a mixed millet/rice cultivation prior to 5000 BP [@Huan2022].

@@

Although the current finding at Guijiabao and Gaoshan support the spread of mixed farming of millet/rice into southwest China around 5000 BP, what their proportions are in local culinary practices during the early time are unclear. For example, the analysis of plant remains and phytolith analysis for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice in neighboring sites, such as Gaoshan and Baodun, Henglanshan, and Baiyangcun in Sichuan and Yunnan [@Hao2022; @Guedes2013]. The analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, but evidence directly reflects food preparation or processing using ceramics still lacks. This study aims to provide direct evidence from food residues in ceramics to investigate culinary practices associated with the cultivation of new crops.

@@

## Samples
We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX). Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and one fragment was from the 2016 season. Those potsherds can be divided into two temporal phases, the early Neolithic period (5000-4500 BP) and the historical period (1000 BP). The common ceramic is cord-marked sand-tempered vessels unearthed at Guijiabao, and their forms include jars, urns, bowls, and basins [@Guijiabao2017a; @Hao2022]. Our samples are mostly reddish-brown jar fragments with medium-size sand temper and some of them have incised cord-marked decoration. Although the original function of those jars is uncertain, they should be supposed to be used for cooking or storage according to the char on the surfaces and their archaeological contexts from a settlement [@Hao2022; @Guijiabao2017a]. This research also attempts to confirm that those ceramics were cooking pots used in the daily life of the local residents.

@@

# Discussion
<!--A brief summary of interesting results-->
Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by ruminant fats during the Neolithic period. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might originate from C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots. The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020]. But we also found a base sample dates to early Neolithic period contains a rich composition of long-chain fatty acids, reflecting a good preservation. We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes. Also, the lack of local modern references for animal fats and plants has impacts on our explanation. However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence at Guijiabao.

@@

<!--discussion about possible ruminant fat origin-->
Ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, accompanied with stone tools for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. But whether horse was part of local food choices in the Neolithic period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. The evidence from Guijuabao also confirms a later presence of horse remains. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats fed on C3 diet based on global references [@Evershed2002]. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016]. A detailed analysis of faunal remains from Guijiabao is still in progress that can provide further information.

We observed a mixture of ruminant fats and terrestrial plants based on a number of Guijiabao potsherds dating to the historical period (ca. 1000 BP). However, we are unable to determine whether there is a shift in diet over time due to our limited sample size. But the overall depleted δ^13^C values for most potsherds suggest C3 plants might be an important diet for local animals and could also be included in the local diet in Yanyuan during the Neolithic transition. This corresponds with the timing when new crops spread into southwest China. Despite of no C4 plant origin in our samples, one potshard implies a source of C4 plant-fed ruminant. In general, the food source detected from our samples comprised mainly of ruminant fats that are more likely from wildlife. This might be related to local ecological zones and environments where wild animals were easier to obtain. Since zooarchaeological analysis and other studies on stone tools are still in progress, we could only rely on preliminary excavation reports with limited details about associated artifacts. A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in food choices during the Neolithic transition.
